Procurement Dynamics: Analyzing Federal Contract Awards to Microsoft

### 2024-12-04 paper/paper.qmd
Final Rearrangement

```{r}
#| label: fig-contract
#| fig-cap: The histogram shows the distribution of performance period, which demonstrates the days between start and end date.
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# create correlation matrix
ggcorrplot(correlation_matrix, lab = TRUE) + 
  theme(
    axis.text.x = element_text(size = 10),  # Adjust x-axis text size
    axis.text.y = element_text(size = 10) # Adjust y-axis text size
  )
```
More information about the data analysis can be found in @sec-result.

@@

## Data

## Model Summary

@@

## Contract Complexity and Cost
The positive association between $ContractDays$ and contract amount suggests that longer contracts are typically more expensive, reflecting the increased scope, complexity, and resource requirements of these agreements. This is particularly relevant for IT-related contracts, which often involve extensive development, maintenance, and support services.

## Negotiation and Efficiency
The negative association between $PhaseDays$ and contract amount indicates that longer preparatory phases may lead to reduced costs, potentially due to more thorough planning, negotiation, or budgetary constraints. This insight can inform government strategies for managing procurement timelines to achieve cost efficiency.

## Sector-Specific Effects 
The significant variation in contract amounts across different contract types (e.g., Consulting, Software) and $buyers$ (e.g., National Defence, Employment and Social Development Canada) underscores the importance of sector-specific factors in procurement dynamics. Departments with higher budgets or critical mandates, such as National Defence, tend to award larger contracts, highlighting their influence in shaping overall procurement trends.

@@

The Bayesian model has low adjusted $R^2$ values. This means the model explains only a small portion (28.9%) of the variance in contract amounts. This suggests that important predictors may be missing. Also, missing or erroneous values in variables like $ContractDays$ or $PhaseDays$ could introduce bias and reduce the reliability of the results.

One limitation of the model is that it may fail to account for certain external factors or unobserved variables such as political factors, budgetary constraints, and departmental priorities. These external factors could lead to omitted variable bias, which also influence procurement decision.

The model assumes a linear relationship between the independent variables and the dependent variable, which oversimplify the dynamics of federal procurement. The model categorizes contract types and buyers into broad groups, which also oversimplify the true diversity of contracts awarded. Future studies could explore non-linear models, such as splines or machine learning approaches, to capture more complex patterns.

There may be issues of endogeneity, where causality runs in both directions between some predictors and the outcome variable. The amount of procurement from a buyer may influence the allocation of future budgets to that department, which in turn affects future contract amounts awarded to Microsoft. This reciprocal relationship could introduce bias in the coefficient estimates if not properly addressed.

@@

This paper provides a foundation for a deeper understanding of federal contract awards to Microsoft, but it also suggests several directions for future exploration. First, the study can expand to include comparisons with other major suppliers. By examining whether Microsoft's procurement trends are unique or reflective of broader patterns across industries. This gives a better understanding on assess whether certain dynamics are systemic within federal contracting practices.

Data quality and consistency issues are also of concern. The anomalies identified in this study, such as incomplete or inconsistent contract dates, highlight the need for improved data reporting standards. These discrepancies can be addressed in collaboration with government agencies to improve the reliability of procurement datasets and ensure that future analyses are more reliable and actionable.

@@

>>>>

@@

More information about data analysis and figures are displayed in @sec-data-visualization.

@@

@@

## Contract Complexity and Cost
The positive association between $ContractDays$ and contract amount suggests that longer contracts tend to be more expensive. This pattern likely reflects the increased scope, complexity, and resource demands of these agreements, which are particularly prominent in IT-related contracts. These contracts often require ongoing development, maintenance, and support, making them more costly over time. Understanding this relationship helps stakeholders anticipate the financial requirements of long-term contracts and make more informed decisions when negotiating the terms of such agreements.

## Negotiation and Efficiency
The negative association between $PhaseDays$ and contract amount suggests that longer planning and negotiation periods may be associated with lower costs. This could be attributed to more thorough upfront planning, strategic negotiation, and budget management, which might lead to better cost control. For procurement agencies, this finding highlights the potential for improving procurement efficiency by streamlining the preparation phase. It also emphasizes the importance of early planning in securing cost-effective contracts, which can ultimately benefit both suppliers and government departments.

## Sector-Specific Effects 
The significant variation in contract amounts across different contract types (e.g., Consulting, Software) and buyers (e.g., National Defence, Employment and Social Development Canada) underscores the importance of sector-specific factors in federal procurement dynamics. Departments with higher budgets or those managing critical projects, such as National Defence, tend to award larger contracts. This observation suggests that the size and nature of the buyer's mandate play a substantial role in shaping procurement patterns. It also indicates that procurement decisions are not solely based on market competitiveness but are influenced by the strategic priorities of the departments involved.

@@

The Bayesian model used in this study has low adjusted $R^2$ values, indicating that it explains only a small portion of the variance in contract amounts (28.9%). This suggests that other important predictors, not captured by the model, may be influencing contract outcomes. The presence of missing or erroneous values in key variables like $ContractDays$ or $PhaseDays$ may introduce bias, further reducing the model's reliability.

One notable limitation of the model is its inability to account for external factors that could significantly affect procurement decisions, such as political influences, departmental priorities, or budgetary constraints. These unobserved variables may introduce omitted variable bias, affecting the model's accuracy. Furthermore, the model assumes linear relationships between the independent variables and the dependent variable, which oversimplifies the complex nature of federal procurement. The categorization of contract types and buyers into broad groups also limits the model's ability to capture the full diversity of contracts awarded.

Endogeneity is another issue that may affect the model's findings. There may be a bidirectional relationship between procurement amounts and buyer behavior—future procurement amounts could influence the allocation of future budgets, which in turn may impact future contracts awarded to Microsoft. This reciprocal relationship, if not properly addressed, can lead to biased coefficient estimates.

@@

This paper provides a foundation for understanding federal procurement contracts awarded to Microsoft but also suggests several directions for future research. One potential extension of this study is to compare Microsoft's procurement trends with those of other major suppliers. By analyzing whether Microsoft's patterns are unique or consistent with broader trends across industries, researchers could better assess whether the observed dynamics are systemic within federal procurement practices.

Additionally, the data quality issues identified—such as incomplete or inconsistent contract dates—highlight the need for improved data reporting standards. These inconsistencies could be addressed through collaboration with government agencies to ensure that future procurement datasets are more accurate and reliable. This would enhance the validity of future analyses and make procurement data more actionable for both policymakers and industry stakeholders.

Finally, exploring more complex models, such as non-linear regression or machine learning approaches, could provide deeper insights into the factors influencing contract amounts. These methods may offer more flexibility in capturing non-linear relationships and uncovering hidden patterns within the data, improving the robustness and predictive power of the analysis.

@@

# Additional Data details {#sec-data-details}

## Graphs {#sec-data-visualization}
```{r}
#| label: fig-contract
#| fig-cap: The histogram shows the distribution of performance period, which demonstrates the days between start and end date.
#| echo: false
#| eval: true
#| warning: false
#| message: false

# create correlation matrix
ggcorrplot(correlation_matrix, lab = TRUE) + 
  theme(
    axis.text.x = element_text(size = 10),  # Adjust x-axis text size
    axis.text.y = element_text(size = 10) # Adjust y-axis text size
  )
```

## Data Visualization Analysis
